---
title: "differential binding cell-SELEX 4th pool"
output: html_notebook
---

Load necessarry packages
```{r, results= "hide"}
library(tidyverse)
library(limma)
library(edgeR)
```

Import the dastasets
```{r, results = 'hide'}

#Sample1 = RCC-MF P4.4, 3 replicates, parallel design 
RCCMF5_url <- "https://www.dropbox.com/s/iq0emuark3frwg5/fastaptamer_enrich_s1.tsv?dl=1"
RCCMF5_file <- tempfile()
download.file(RCCMF5_url, RCCMF5_file, mode = "wb")
sample1 <- read_tsv(RCCMF5_file)
#Sample2 = RC-124 P4.4, 3 replicates, parallel design
RC1245_url <- "https://www.dropbox.com/s/64y4zq50106s0zo/fastaptamer_enrich_s2.tsv?dl=1"
RC1245_file <- tempfile()
download.file(RC1245_url, RC1245_file, mode = "wb")
sample2 <- read_tsv(RC1245_file)
```

COmbine both datasets.

Sample1 = RCC_MF binding P4 sequences, 3 replicates

Sample2 = RC_124 binding P4 sequences, 3 replicates

```{r}
apta_input_1<- full_join(sample2, sample1, by = "Sequence")
apta_input_1

```

Select the columns that are necessary and replace NAs with 0.

```{r}
apta_input_2 <- apta_input_1 %>% 
  select('Reads (x).x', 'Reads (y).x', 'Reads (z).x', 'Reads (x).y', 'Reads (y).y', 'Reads (z).y') %>% 
  replace_na(list('Reads (x).x' = 0, 'Reads (y).x'=0, 'Reads (z).x'=0, 'Reads (x).y'=0, 'Reads (y).y'=0, 'Reads (z).y'=0)) %>% 
  sapply(as.numeric)
```

Remove the colnames

```{r}
colnames(apta_input_2) <- NULL
```

Create grouping vector - 3 replicates for each group.
```{r}
group <- factor(c(1,1,1,2,2,2))
```

Create DGEList
```{r}
apta_input_3 <- DGEList(counts = apta_input_2, group = group)
apta_input_3
```

Calculate counts per million and view summary

```{r}
apta_input_3_cpm <- cpm(apta_input_3)
summary(apta_input_3$counts)
```

Filter out sequences that appear in less than 2 times or that have less than 2 reads per million 

```{r}
keep <- rowSums(cpm(apta_input_3)>2) >= 2
apta_input_3_filt <- apta_input_3[keep, , keep.lib.sizes=FALSE]
apta_input_3_filt
summary(apta_input_3_filt$counts)
```


Calculate normalization factors

```{r}
apta_input_4_filt_norm <- calcNormFactors(apta_input_3_filt)
apta_input_4_filt_norm
nrow(apta_input_4_filt_norm)
```  

Create a model matrix for experiment

```{r}
design_apta <- model.matrix(~group)
```

Plot MDS

```{r}
plotMDS(apta_input_4_filt_norm)
```

Estimate dispersion

```{r}
#apta_de <- estimateDisp(apta_input_4_filt_norm, design_apta)
apta_de <- estimateGLMCommonDisp(apta_input_4_filt_norm, design_apta)
apta_de <- estimateGLMTrendedDisp(apta_de, design_apta)
apta_de <- estimateGLMTagwiseDisp(apta_de, design_apta)
apta_de$common.dispersion
```

Plot BCV

```{r}
plotBCV(apta_de)
```



```{r}
fit <- glmFit(apta_de, design_apta)
lrt <- glmLRT(fit, coef = 2)
lrt$table
top_apta <- topTags(lrt,n = 2000L, sort.by = "logFC", adjust.method = "fdr", p.value = 0.0001)
top_apta_table <- top_apta$table
top_apta_table
```



```{r}
dbapta <- decideTestsDGE(lrt, adjust.method = "fdr", p.value=0.0001, lfc = 2)
dbapta <- rownames(lrt)[as.logical(dbapta)]
length(dbapta)
plotSmear(lrt, de.tags=dbapta)+abline(h = c(-2, 0, 2), col = c("dodgerblue","yellow","dodgerblue"), lty=2)
```



