---
title: "differential binding cell-SELEX combined with enrichment analysis"
output: html_notebook
---
Load necessary packages


```{r, results= "hide"}
library(tidyverse)
library(limma)
library(edgeR)
```


Import the dastasets

```{r, results = 'hide'}
#Smaple1 = RCC-MF P4.4, 3 replicates, parallel design 
RCCMF5_url <- "https://www.dropbox.com/s/iq0emuark3frwg5/fastaptamer_enrich_s1.tsv?dl=1"
RCCMF5_file <- tempfile()
download.file(RCCMF5_url, RCCMF5_file, mode = "wb")
sample1 <- read_tsv(RCCMF5_file)
#Sample3 = RCC-MF P11, 3 replicates, parallel design
sample3_url <- "https://www.dropbox.com/s/f2j8fmxmbbqmgub/fastaptamer_enrich_s3.tsv?dl=1"
sample3_file <- tempfile()
download.file(sample3_url, sample3_file, mode = "wb")
sample3 <- read_tsv(sample3_file)
#Sample4 = RC-124 P11, 3 replicates, parallel design
sample4_url <- "https://www.dropbox.com/s/gfdgl3y9210qymz/fastaptamer_enrich_s4.tsv?dl=1"
sample4_file <- tempfile()
download.file(sample4_url, sample4_file, mode = "wb")
sample4 <- read_tsv(sample4_file)
```

COmbine both datasets.

Sample3 = RCC_MF binding P11 sequences, 3 replicates

Sample4 = RC_124 binding P11 sequences, 3 replicates

```{r}
apta_input_1<- full_join(sample4, sample3, by = "Sequence")
apta_input_1

```

Select the columns that are necessary and replace NAs with 0.

```{r}
apta_input_2 <- apta_input_1 %>% 
  select('Reads (x).x', 'Reads (y).x', 'Reads (z).x', 'Reads (x).y', 'Reads (y).y', 'Reads (z).y') %>% 
  replace_na(list('Reads (x).x' = 0, 'Reads (y).x'=0, 'Reads (z).x'=0, 'Reads (x).y'=0, 'Reads (y).y'=0, 'Reads (z).y'=0))
apta_input_2
```

Remove the colnames

```{r}
colnames(apta_input_2) <- NULL
apta_input_2
```

Create grouping vector - 3 replicates for each group

```{r}
group <- factor(c(1,1,1,2,2,2))
```

Create DGEList

```{r}
apta_input_3 <- DGEList(counts = apta_input_2, group = group)
apta_input_3
```

Calculate counts per million and view summary

```{r}
apta_input_3_cpm <- cpm(apta_input_3)
summary(apta_input_3$counts)
```

Filter out sequences that appear in less than 2 times or that have less than 2 reads per million

```{r}
keep <- rowSums(cpm(apta_input_3)>2) >= 2
apta_input_3_filt <- apta_input_3[keep, , keep.lib.sizes=FALSE]
summary(apta_input_3_filt$counts)
```


Calculate normalization factors

```{r}
apta_input_4_filt_norm <- calcNormFactors(apta_input_3_filt)
apta_input_4_filt_norm
nrow(apta_input_4_filt_norm)
```  

Create a model matrix for experiment

```{r}
design_apta <- model.matrix(~group)
```

Plot MDS

```{r}
plotMDS(apta_input_4_filt_norm)
```

Estimate dispersion

```{r}
#apta_de <- estimateDisp(apta_input_4_filt_norm, design_apta)
apta_de <- estimateGLMCommonDisp(apta_input_4_filt_norm, design_apta)
apta_de <- estimateGLMTrendedDisp(apta_de, design_apta)
apta_de <- estimateGLMTagwiseDisp(apta_de, design_apta)
apta_de$common.dispersion
```

Plot BCV

```{r}
plotBCV(apta_de)
```

Fit negative binomial generalized log-linear model to read counts.



```{r}
fit <- glmFit(apta_de, design_apta)
lrt <- glmLRT(fit, coef = 2)

top_apta <- topTags(lrt,n = 2000L, sort.by = "logFC", adjust.method = "fdr", p.value = 0.0001)
top_apta_table <- top_apta$table

#table of statistically significantly differentially bound aptamers WITHOUT filtering by logFC>2

length(top_apta_table$logFC)

```



```{r}
dbapta <- decideTestsDGE(lrt, adjust.method = "fdr", p.value=0.0001, lfc = 2)

##table of statistically significantly differentially bound aptamers WITH filtering by logFC>2
dbapta <- rownames(lrt)[as.logical(dbapta)]
length(dbapta)
dbapta
```
Ggplot2

```{r}
dbapta_sign_only<- lrt$table[dbapta,]
length(dbapta_sign_only$logFC)
p = ggplot(dbapta_sign_only, aes(x =dbapta_sign_only$logFC, y = dbapta_sign_only$logCPM)) + geom_point(aes(colour="red")) + theme_bw() + geom_vline(xintercept = c(-2, 2))

p
```


Identify the sequences that are differentially bound:

1. Identify sequences based on returned deGenes row numbers;

2. Filter out results that have -logFC < 2;

3. Match the data in original combined file

4. Return the sequences

```{r}
#total number of statistically significantly bound aptamer sequences
#should be 198
length(dbapta_sign_only$logFC)

#number of statistically significantly differentially bound aptamer sequences with log>2
#in other words - sequences that bind RCC-MF
#shuld be 17
top_apta_log2_filtered_RCC <- dbapta_sign_only %>% 
  rownames_to_column() %>% 
  filter(dbapta_sign_only$logFC > 2)
length(top_apta_log2_filtered_RCC$rowname)


#number of statistically significantly differentially bound aptamer sequences with log<-2
#in other words - sequences that bind RC-124
#shuld be 178
top_apta_log2_filtered_RC124<-dbapta_sign_only %>% 
  rownames_to_column() %>% 
  filter(dbapta_sign_only$logFC < -2)
length(top_apta_log2_filtered_RC124$rowname)


#significant sequences that bind RCC-MF
top_apt_matched_RCC<- apta_input_1[top_apta_log2_filtered_RCC$rowname,]
top_apta_matched_reads_RCC <- top_apt_matched_RCC[,c(1,4, 7,10, 17, 20, 23)]
top_apta_matched_reads_RCC

#significant sequences that bind RC-124
top_apt_matched_RC124<- apta_input_1[top_apta_log2_filtered_RC124$rowname,]
top_apta_matched_reads_RC124 <- top_apt_matched_RC124[,c(1,4, 7,10, 17, 20, 23)]
top_apta_matched_reads_RC124

```

After differential binding data have been aquired, combine this information with classical enrichment analysis
First, introduce this fuction to combine all replicates from samples 1 and 3 into 1 table per samples 

```{r}
aptamer_analysis <- function(enrich_tsv) {
  a <- enrich_tsv %>% 
    select(Sequence = "Sequence", RPM_3 = "RPM (x)", RPM_2 = "RPM (y)", RPM_1 = "RPM (z)") %>%
    gather("RPPM", "RPM", 2:4, na.rm = TRUE)
  a$RPM <- as.numeric(a$RPM)
  a_2 <- a %>% 
    group_by(Sequence) %>% 
    summarise(meanRPM = mean(RPM), n = n(), sd = sd(RPM), RSD = sd(RPM)/mean(RPM)*100)
}
```

Apply the function to RCC-MF binding after 4th cycle of selection, 11th cycle of selection and binding to RC-124 after 11th cycle

```{r}
sample_1_enrich <- aptamer_analysis(sample1)
sample_3_enrich <- aptamer_analysis(sample3)
sample_4_enrich <- aptamer_analysis(sample4)

```

Combine both files, calculate the logFC enrichment and filter according to the value that you choose. In this case logFC value of 6 was used which means that the sequences were enriched at least 64 times

```{r}
comb_13 <- full_join(x = sample_1_enrich, y = sample_3_enrich, by = "Sequence")
comb_13$log2_change <- log2(comb_13$meanRPM.y/comb_13$meanRPM.x)
comb_13_log2_filtered <- comb_13 %>% 
  filter(log2_change > 6)

nrow(comb_13_log2_filtered)
```

At the end we combine the sequences that were differentially bound to sequences that were enriched to identify the most relevant aptamer sequences.

Some of the sequences that were differentially boun did not appear in enriched dataset with applied logFC value.

```{r}
#match sequences from differential binding to RCC-MF to sequences from enrichment analysis
#should be 2
match_to_diff_RCC <- semi_join(comb_13_log2_filtered, top_apta_matched_reads_RCC, by = "Sequence")
match_to_diff_RCC
nrow(match_to_diff_RCC)
```

```{r}
#match sequences that are both enriched and differentially bound to RCC-MF cells to sequences that are in RC-124 dataset to see how different is the abundance between the pools
#should be 2
match_to_RC124_p11 <- semi_join(sample_4_enrich, match_to_diff_RCC, by="Sequence")
match_to_RC124_p11
```


```{r}
#calculate the log2 differential binding value between RCC-MF and RC-124 and log2 enrichment value from 4th cycle to 11th cycle to RCC-MF cells
match_to_diff_RCC$testlog2toRC124<- log2(match_to_diff_RCC$meanRPM.y/match_to_RC124_p11$meanRPM)
match_to_diff_RCC %>% 
  select("Sequence", "log2_enrichment" = "log2_change", "log2FC_compared_to_RC124" = "testlog2toRC124")

```

Make a plot with all sequences

```{r}
plotSmear(lrt, de.tags=dbapta)+abline(h = c(-2, 0, 2), col = c("dodgerblue","yellow","dodgerblue"), lty=2)
```
